On the number of tiles visited by a\\ line segment on a rectangular grid

### 2023-04-18 paper.tex
Wording
Minor adjustments

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\cond}{\,|\,} % less spacing than \mid
\newcommand{\diff}{\mathrm d}
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\DeclareMathOperator{\E}{E}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gives infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\funta}{\varphi} % function that gives average number of touched tiles for real-valued length
%\newcommand{\funca}{\psi} % function that gives average number of grid line crossings for real-valued length
\newcommand{\probmax}{\rho} % function that gives the probability of achieving the maximum number of touched tiles in the square case for real-valued length
\newcommand{\ras}{\sigma}
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\genfun}{q}
\newcommand{\genvar}{s}
\newcommand{\roundterm}{r}
\newcommand{\mss}{M}
%\newcommand{\nextrat}{K}
\newcommand{\funcirc}{\Omega}
\newcommand{\funcircesc}{\bar \Omega}
\newcommand{\fdp}{p}

@@

\begin{theorem}
\label{theo: funt, form}
For $a, b \in \mathbb R^+$, $a \geq b$ and $\len \in \mathbb R^+$,
\begin{equation}
\label{eq: theo: funt, form; funt}
\funt(\len) = \isoli+\jsoli-1
\end{equation}
with
\begin{align}
\label{eq: theo: funt, form; i}
\isoli &= \left\lceil \frac 3 2 + \frac b a \Re \sqrt{\frac{\len^2}{a^2+b^2}-\frac 1 4} \right\rceil, \\
\label{eq: theo: funt, form; j}
\jsoli &= \left\lceil 1 + \frac{\sqrt{\len^2-(\isoli-2)^2a^2}}{b} \right\rceil.
\end{align}
The function $\funt$ is piecewise constant and left-continuous, with unit-height jumps. A jump occurs at $\len$ if and only if $\len = \funl(\tiles)$ for some $\tiles \in \mathbb N$, $\tiles \geq 4$; and then $\funt(\len) = \tiles-1$, $\lim_{h \rightarrow 0+} \funt(\len+h) = \tiles$.
\end{theorem}

@@

Consider an arbitrary $\len$ such that for some $\tiles \in \mathbb N$, $\tiles \geq 4$
\begin{equation}
\label{eq: theo funt, form: prop 1}
\funl(\tiles)=\len.
\end{equation}
To see that $\funt$ has a jump at $\len$, assume for the sake of contradiction that $\funt$ is continuous at that point. Therefore $\funt$ is constant on an interval containing that point, which implies that $\funt(\len-\epsilon) = \tiles = \funt(\len+\epsilon)$ for some $\epsilon > 0$. This means that there exists a segment with length $\len-\epsilon$ that visits $\tiles$ tiles, and thus $\funl(\tiles) \leq \len-\epsilon < \len$, in contradiction with \eqref{eq: theo funt, form: prop 1}. Therefore $\funt$ is discontinuous (from the right) at $\len$. By definition of $\funl$, from \eqref{eq: theo funt, form: prop 1} it follows that
\begin{equation}
\label{eq: theo funt, form: prop 2}
\funt(\len) < \tiles
\end{equation}
and there exists $\epsilon > 0$ such that $\funt(\len+h) = \tiles$ for $0 < h < \epsilon$. This implies that
\begin{equation}
\label{eq: theo funt, form: prop 3}
\lim_{h \rightarrow 0+} \funt(\len+h) = \tiles,
\end{equation}
that is, $\funt$ has a jump at $\len$. In addition, since the jump has unit height, it stems from \eqref{eq: theo funt, form: prop 2} and \eqref{eq: theo funt, form: prop 3}  that $\funt(\len) = \tiles-1$.

@@

For $0 \leq \genvar \leq 1$, let
\begin{equation}
\label{eq: f}
f(\genvar) = \int_0^\genvar \arccos z \, \diff z = \genvar \arccos \genvar - \sqrt{1-\genvar^2} + 1.
\end{equation}

@@

The probability of visiting the maximum number of tiles for a rectangular grid is difficult to compute. The reason is analogous to the irregularity of the optimal $(i,j)$ pairs which was discussed in \S\ref{part: fund results}. In a square grid, however, the problem is more tractable.

@@

For $u, v \in \mathbb N$, $\genvar \in \mathbb R^+$, let
\begin{equation}
\label{eq: g def}
g(\genvar, u, v) = \frac 2 \pi \int_{\arcsin \frac v {\genvar}}^{\arccos \frac u {\genvar}} \left( \genvar \cos \theta - u \right) \left( \genvar \sin \theta - v \right) \, \diff \theta.
\end{equation}
Computing this integral and using the identities $\cos^2 \alpha = 1-\sin^2 \alpha$ and $\cos(2\alpha) = 2\cos^2\alpha-1 = 1 - 2\sin^2\alpha$,
\begin{equation}
\label{eq: g fin}
\begin{split}
g(\genvar, u, v) &= \frac 1 {\pi} \biggl(
2\left(\arccos\frac{u}{\genvar}-\arcsin\frac{v}{\genvar}\right) u v + \genvar^2 + u^2 + v^2 \\
& \quad - 2 u \sqrt{\genvar^2-v^2} - 2 v \sqrt{\genvar^2-u^2} \biggr).
\end{split}
\end{equation}
Taking into account that $\arcsin x = \pi/2 - \arccos x$, it follows from \eqref{eq: g fin} that $g(\genvar, u, v) = g(\genvar, v, u)$.

@@

Let $\tilde y_{\tiles,n}$ denote the $y$ coordinate of the chord for $x = x_{\tiles,n}$, as shown in Figure~\ref{fig: probmax_asint_detail}. By construction $|m_{\tiles,n}|<1$. Consider $a_{\tiles,n}$ and $b_{\tiles,n}$ as defined in the figure. It can be seen that
\begin{equation}
\frac{\tilde y_{\tiles,n}-a_{\tiles,n}/\sqrt{2}}{a_{\tiles,n}/\sqrt{2}} = m_{\tiles,n},
\end{equation}
which gives $a_{\tiles,n} = \sqrt{2}\tilde y_{\tiles,n}/(1+m_{\tiles,n})$. Likewise, $b_{\tiles,n} = \sqrt{2}\tilde y_{\tiles,n}/(1-m_{\tiles,n})$. With the upper corner of the shaded rectangle uniformly distributed along the chord, its average area $\tilde A_{\tiles,n}$ is easily computed in terms of $a_{\tiles,n}$ and $b_{\tiles,n}$, and therefore of $\tilde y_{\tiles,n}$ and $m_{\tiles,n}$:
\begin{equation}
\label{eq: tilde A y}
\tilde A_{\tiles,n} = a_{\tiles,n}b_{\tiles,n} \int_0^1 \eta(1-\eta)\, \diff \eta = \frac{a_{\tiles,n}b_{\tiles,n}} 6 = \frac{\tilde y_{\tiles,n}^2}{3(1-m_{\tiles,n}^2)}.
\end{equation}
Similarly,
\begin{equation}
\label{eq: tilde c y}
\tilde c_{\tiles,n} = \sqrt{a_{\tiles,n}^2+b_{\tiles,n}^2} = \frac {2 \tilde y_{\tiles,n} \sqrt{1+m_{\tiles,n}^2}} {1-m_{\tiles,n}^2}.
\end{equation}
Substituting \eqref{eq: tilde A y} and \eqref{eq: tilde c y} into \eqref{eq: P i j asympt 3 bis},
\begin{equation}
\label{eq: P i j asympt 4}
P_{\tiles,n}
= \frac{4\sqrt{2} \sqrt{1+m_{\tiles,n}^2}} {3\pi(1-m_{\tiles,n}^2)^2}\, \tilde y_{\tiles,n}^3 + E_{\tiles, n}.
\end{equation}
The inequality \eqref{eq: asympt slope bound} applies in particular to $|m_{\tiles,n}|$, and gives $m_{\tiles,n}^2 < 9/(2(\tiles-3))$. For simplicity, assume $\tiles \geq 8$ so that this bound is less than $1$. Then, making use of \eqref{eq: sqrt asympt}, the term depending on $m_{\tiles,n}^2$ in \eqref{eq: P i j asympt 4} satisfies
\begin{equation}
\label{eq: P i j asympt 4 bis}
\frac{\sqrt{1+m_{\tiles, n}^2}} {(1-m_{\tiles, n}^2)^2} \leq \frac{(\tiles-3)^2}{(\tiles-15/2)^2} \left(1+\frac {9/4} {\tiles-3}\right) = 1 + \frac{9(5\tiles - 24)}{(2\tiles-15)^2}.
\end{equation}
From \eqref{eq: h bound},
\begin{equation}
\label{eq: P i j asympt 4 ter}
\tilde y_{\tiles,n} < y_{\tiles,n} \leq h_\tiles \leq \frac{3\sqrt{2}} 4.
\end{equation}
Substituting \eqref{eq: P i j asympt 4 bis} into \eqref{eq: P i j asympt 4} and taking into account \eqref{eq: P i j asympt 4 ter},
\begin{equation}
\label{eq: P i j asympt 5}
P_{\tiles,n} = \frac{4\sqrt{2}} {3\pi} \tilde y_{\tiles,n}^3 + E_{\tiles,n} + E'_{\tiles,n}
\end{equation}
with $|E'_{\tiles,n}| < e'(t)$, where $e'(t)$ tends to $0$ as $t \rightarrow \infty$.

>>>>

\newcommand{\mr}{\mathrm}
\newcommand{\st}{\mid}%{\,|\,}
\newcommand{\cond}{\,|\,} % less spacing than \mid
\newcommand{\diff}{\mathrm d}
\let\Re\relax % remove the command...
\DeclareMathOperator{\Re}{Re} % ...to define it differently
\let\Im\relax
\DeclareMathOperator{\Im}{Im}
\DeclareMathOperator{\E}{E}
\newcommand{\funt}{\tau} % function that gives maximum number of touched tiles for real-valued length
\newcommand{\funl}{\lambda} % function that gives infimum real-valued length for a given number of touched tiles
\newcommand{\funti}{T} % function that gives maximum number of touched tiles for integer-valued length
\newcommand{\funli}{\Lambda} % function that gives minimum integer-valued length for a given number of touched tiles
\newcommand{\funta}{\varphi} % function that gives average number of touched tiles for real-valued length
%\newcommand{\funca}{\psi} % function that gives average number of grid line crossings for real-valued length
\newcommand{\probmax}{\rho} % function that gives the probability of achieving the maximum number of touched tiles in the square case for real-valued length
\newcommand{\ras}{\sigma}
\newcommand{\len}{\ell} % length variable, real-valued
\newcommand{\leni}{\ell} % length variable, integer-valued. Same notation, but I keep the different command just in case
\newcommand{\tiles}{t} % number of tiles (integer-valued)
\newcommand{\isolr}{i^+}
\newcommand{\jsolr}{j^+}
\newcommand{\isoli}{i^\ast}
\newcommand{\jsoli}{j^\ast}
\newcommand{\genfun}{q}
\newcommand{\genvar}{s}
\newcommand{\roundterm}{r}
\newcommand{\mss}{M}
%\newcommand{\nextrat}{K}
\newcommand{\funcirc}{\Omega}
\newcommand{\funcircesc}{\bar \Omega}
\newcommand{\fdp}{p}
\newcommand{\lra}{\alpha}
\newcommand{\lrb}{\beta}

@@

\begin{theorem}
\label{theo: funt, form}
For $a, b \in \mathbb R^+$, $a \geq b$ and $\len \in \mathbb R^+$,
\begin{equation}
\label{eq: theo: funt, form; funt}
\funt(\len) = \isoli+\jsoli-1
\end{equation}
with
\begin{align}
\label{eq: theo: funt, form; i}
\isoli &= \left\lceil \frac 3 2 + \frac b a \Re \sqrt{\frac{\len^2}{a^2+b^2}-\frac 1 4} \right\rceil, \\
\label{eq: theo: funt, form; j}
\jsoli &= \left\lceil 1 + \frac{\sqrt{\len^2-(\isoli-2)^2a^2}}{b} \right\rceil.
\end{align}
The function $\funt$ is piecewise constant and left-continuous, with unit-height jumps. A jump occurs at $\len$ if and only if $\len = \funl(\tiles)$ for some $\tiles \in \mathbb N$, $\tiles \geq 4$; and then $\funt(\len) = \tiles-1$, $\lim_{\delta \rightarrow 0+} \funt(\len+\delta) = \tiles$.
\end{theorem}

@@

Consider an arbitrary $\len$ such that for some $\tiles \in \mathbb N$, $\tiles \geq 4$
\begin{equation}
\label{eq: theo funt, form: prop 1}
\funl(\tiles)=\len.
\end{equation}
To see that $\funt$ has a jump at $\len$, assume for the sake of contradiction that $\funt$ is continuous at that point. Therefore $\funt$ is constant on an interval containing that point, which implies that $\funt(\len-\epsilon) = \tiles = \funt(\len+\epsilon)$ for some $\epsilon > 0$. This means that there exists a segment with length $\len-\epsilon$ that visits $\tiles$ tiles, and thus $\funl(\tiles) \leq \len-\epsilon < \len$, in contradiction with \eqref{eq: theo funt, form: prop 1}. Therefore $\funt$ is discontinuous (from the right) at $\len$. By definition of $\funl$, from \eqref{eq: theo funt, form: prop 1} it follows that
\begin{equation}
\label{eq: theo funt, form: prop 2}
\funt(\len) < \tiles
\end{equation}
and there exists $\epsilon > 0$ such that $\funt(\len+\delta) = \tiles$ for $0 < \delta < \epsilon$. This implies that
\begin{equation}
\label{eq: theo funt, form: prop 3}
\lim_{\delta \rightarrow 0+} \funt(\len+\delta) = \tiles,
\end{equation}
that is, $\funt$ has a jump at $\len$. In addition, since the jump has unit height, it stems from \eqref{eq: theo funt, form: prop 2} and \eqref{eq: theo funt, form: prop 3}  that $\funt(\len) = \tiles-1$.

@@

For $0 \leq \genvar \leq 1$, let
\begin{equation}
\label{eq: f}
f(\genvar) = \int_0^\genvar \arccos x \, \diff x = \genvar \arccos \genvar - \sqrt{1-\genvar^2} + 1.
\end{equation}

@@

The probability of visiting the maximum number of tiles for a rectangular grid is difficult to compute. The reason is the irregularity of the relevant $(i,j)$ pairs for general $a$, $b$, analogous to that observed in \S\ref{part: fund results} for the pairs $(i,j)$ that maximize the number of visited tiles. In a square grid, however, the problem is more tractable.

@@

For $u, v \in \mathbb N$, $\genvar \in \mathbb R^+$, let
\begin{equation}
\label{eq: g def}
g(\genvar, u, v) = \frac 2 \pi \int_{\arcsin \frac v {\genvar}}^{\arccos \frac u {\genvar}} \left( \genvar \cos \theta - u \right) \left( \genvar \sin \theta - v \right) \, \diff \theta.
\end{equation}
Computing this integral and using the identities $\cos^2 z = 1-\sin^2 z$ and $\cos(2z) = 2\cos^2z-1 = 1 - 2\sin^2z$,
\begin{equation}
\label{eq: g fin}
\begin{split}
g(\genvar, u, v) &= \frac 1 {\pi} \biggl(
2\left(\arccos\frac{u}{\genvar}-\arcsin\frac{v}{\genvar}\right) u v + \genvar^2 + u^2 + v^2 \\
& \quad - 2 u \sqrt{\genvar^2-v^2} - 2 v \sqrt{\genvar^2-u^2} \biggr).
\end{split}
\end{equation}
Taking into account that $\arcsin x = \pi/2 - \arccos x$, it follows from \eqref{eq: g fin} that $g(\genvar, u, v) = g(\genvar, v, u)$.

@@

Let $\tilde y_{\tiles,n}$ denote the $y$ coordinate of the chord for $x = x_{\tiles,n}$, as shown in Figure~\ref{fig: probmax_asint_detail}. By construction $|m_{\tiles,n}|<1$. Consider the lengths $\lra_{\tiles,n}$ and $\lrb_{\tiles,n}$ defined in the figure. Then
\begin{equation}
\frac{\tilde y_{\tiles,n}-\lra_{\tiles,n}/\sqrt{2}}{\lra_{\tiles,n}/\sqrt{2}} = m_{\tiles,n},
\end{equation}
which gives $\lra_{\tiles,n} = \sqrt{2}\tilde y_{\tiles,n}/(1+m_{\tiles,n})$. Likewise, $\lrb_{\tiles,n} = \sqrt{2}\tilde y_{\tiles,n}/(1-m_{\tiles,n})$. With the upper corner of the shaded rectangle uniformly distributed along the chord, its average area $\tilde A_{\tiles,n}$ is easily computed in terms of $\lra_{\tiles,n}$ and $\lrb_{\tiles,n}$, and therefore of $\tilde y_{\tiles,n}$ and $m_{\tiles,n}$:
\begin{equation}
\label{eq: tilde A y}
\tilde A_{\tiles,n} = \lra_{\tiles,n}\lrb_{\tiles,n} \int_0^1 \eta(1-\eta)\, \diff \eta = \frac{\lra_{\tiles,n}\lrb_{\tiles,n}} 6 = \frac{\tilde y_{\tiles,n}^2}{3(1-m_{\tiles,n}^2)}.
\end{equation}
Similarly,
\begin{equation}
\label{eq: tilde c y}
\tilde c_{\tiles,n} = \sqrt{\lra_{\tiles,n}^2+\lrb_{\tiles,n}^2} = \frac {2 \tilde y_{\tiles,n} \sqrt{1+m_{\tiles,n}^2}} {1-m_{\tiles,n}^2}.
\end{equation}
Substituting \eqref{eq: tilde A y} and \eqref{eq: tilde c y} into \eqref{eq: P i j asympt 3 bis},
\begin{equation}
\label{eq: P i j asympt 4}
P_{\tiles,n}
= \frac{4\sqrt{2} \sqrt{1+m_{\tiles,n}^2}} {3\pi(1-m_{\tiles,n}^2)^2}\, \tilde y_{\tiles,n}^3 + E_{\tiles, n}.
\end{equation}
The inequality \eqref{eq: asympt slope bound} applies in particular to $|m_{\tiles,n}|$, and gives $m_{\tiles,n}^2 < 9/(2(\tiles-3))$. For simplicity, assume $\tiles \geq 8$ so that this bound is less than $1$. Then, making use of \eqref{eq: sqrt asympt}, the term depending on $m_{\tiles,n}^2$ in \eqref{eq: P i j asympt 4} satisfies
\begin{equation}
\label{eq: P i j asympt 4 bis}
\frac{\sqrt{1+m_{\tiles, n}^2}} {(1-m_{\tiles, n}^2)^2} \leq \frac{(\tiles-3)^2}{(\tiles-15/2)^2} \left(1+\frac {9/4} {\tiles-3}\right) = 1 + \frac{9(5\tiles - 24)}{(2\tiles-15)^2}.
\end{equation}
From \eqref{eq: h bound},
\begin{equation}
\label{eq: P i j asympt 4 ter}
\tilde y_{\tiles,n} < y_{\tiles,n} \leq h_\tiles \leq \frac{3\sqrt{2}} 4.
\end{equation}
Substituting \eqref{eq: P i j asympt 4 bis} into \eqref{eq: P i j asympt 4} and taking into account \eqref{eq: P i j asympt 4 ter},
\begin{equation}
\label{eq: P i j asympt 5}
P_{\tiles,n} = \frac{4\sqrt{2}} {3\pi} \tilde y_{\tiles,n}^3 + E_{\tiles,n} + E'_{\tiles,n}
\end{equation}
with $|E'_{\tiles,n}| < e'(t)$, where $e'(t)$ tends to $0$ as $t \rightarrow \infty$.
